Labor Markets and Technological Change: Evidence from Electronic Health Records

### 2024-03-25 Paper/PaperEdit.tex

Two key assumptions underlie the main analysis. First, I assume that hospitalists are constrained to use the technology implemented in their work setting. Second, I assume the decision for hospitals to implement EHRs is exogenous to individual hospitalists, and uncorrelated with other major changes to the hospitalist work environment. I mitigate concerns of the latter by analyzing whether other observable hospital changes are correlated with EHR adoption, and find no evidence that this could be biasing my results. Additionally, my main findings are not sensitive to a plethora of robustness checks. The results suggest that a surge of hospitalists left clinical settings altogether, and even more that shifted work away from the hospitals where the EHR was deployed and towards outpatient/office settings. Thus, EHR implementation may have intensified access to care issues for hospitalized patients. This unintended consequence of incentivizing rapid technology adoption highlights the need for careful consideration of technologies and policies that might negatively affect work environment.

@@

The Medicare Data on Provider Practice and Specialty (MD-PPAS) is a database of physicians which details claim counts, physician specialty, and practice location from 2009-2017. Variables included are physician NPI, primary and secondary specialties, patient count and total billing counts, and fraction of patients seen in different settings (hospital, office/outpatient, emergency room, etc.). I only include physicians classified as hospitalists\footnote{Physician who self-identifies as hospitalist or has 70\% of patients in inpatient setting. I cannot distinguish between types of hospitalists as the data source manipulates this definition based on percent of patients seen in a hospital, so these can be formal hospitalists or internists.}, internal medicine, general practice or family practice in at least one year. I drop physicians with less than 70\% of patients in a hospital setting in all years, as the focus is on those who are affected by a hospital's decision to adopt (I investigate robustness to this threshold in Appendix \ref{sec:changes}). I use this data to construct the dependent variables for my analysis.

@@

I aggregate the data to the hospitalist level. I drop those who are not treated by 2015 since I do not observe whether they become treated in 2016 or 2017.\footnote{In Appendix \ref{sec:changes} I limit years to 2009-2015 and include never-treated. The results are robust to this specification choice.} That is, the data contains no units which are ``never-treated". Finally, I include a physician's graduation year from Physician Compare (provided by CMS). I limit to hospitalists who graduated before 2005, as anyone who graduated medical school after that will be finishing residency during the span of the data and will exhibit labor market changing behavior which could be correlated to EHR exposure purely because of switching work settings during a time of rapid EHR implementation.There are various thresholds used in the construction of this final data, including how many shared patients constitutes a physician-hospital relationship. The details of such thresholds are explained in Appendix \ref{app:data}. To ensure that my results  are not sensitive to various choices of the thresholds, I estimate the effect of EHR exposure on all outcomes under different combinations of data limitations. The specification charts with these results are shown in Appendix \ref{sec:chart}, and I discuss each in conjunction with the main results.

@@

As discussed in Section \ref{sec:worksetting}, hospitalists could be switching to work settings either because of their own preferences to avoid technology, or because EHRs lead hospitals to re-allocate workers to hospital-owned outpatient settings. While I do not observe individual physician contract type with each hospital, I do observe hospital level information about the nature of aggregate physician relationships. To investigate the potential mechanism of hospital allocation, I present a heterogeneity analysis for all outcomes broken down by level of integration the physician is exposed to in their connected hospitals. From least integrated (contracts with multiple facilities) to most integrated (full salaried hospitalists), results are broken down by Independent Practice Association (IPA), Open Physician-Hospital Organization (OPHO), Closed Physician-Hospital Organization (CPHO), and Integrated Salary Model (ISM). In practice, hospitals can only re-allocate hospitalists to outpatient facilities if the hospitalist is heavily integrated. However, estimates do not differ between the types of integration relationships, indicating that the results are not driven entirely by such re-allocation.

\begin{figure}
    \centering
    \caption{Effect of EHR Exposure on Outcomes by Integration Level}
    \includegraphics[width=\textwidth]{Objects/heterog_plot_int.pdf}
    \label{fig:het_int}
\end{figure}

Additionally, I investigate whether this finding is potentially due EHRs changing the type of billing performed by hospitalists. For example, maybe they are doing the same services in the same place but billing the differently. I do not find this to be the case. This analysis can be found in Appendix

@@

While I am investigating the supply side factors of this outcome, patient count is also driven by demand for health care. A major change in the U.S. health care system, the Affordable Care Act (ACA) was passed in 2010 and took effect in 2014. If my results are driven primarily by the group of hospitals who implemented EHRs in 2014, one might be concerned that the legislation may be driving the increased patient count. The ACA had potentially opposing effects on Medicare patients: they may be crowded out by an increased demand in Medicaid patients, or the small changes to Medicare may have led to an increase in utilization from the Medicare population. A recent study found no negative spillovers to the Medicare population due to the ACA (\cite{carey2020impact}), suggesting that demand change should not drive my results. Additionally, I present average group time treatment effects for each treated group in Figure \ref{fig:patientgroup} of the appendix. Each group shows the same pattern of an increase in patient count after exposure, suggesting that the groups affected by the ACA are not driving the results.

EHR exposure affects which hospitalists are included in this analysis (since I drop those who exit or move), indicating a non-random attrition problem. That is, if productive or tech savvy hospitalists are the ones who remain in the same hospital after EHR exposure, the estimates are biased. The proportion of those who are not included in this analysis is 18\% in the control group and 36\% in the treated group. To investigate whether my results are robust to this issue, I construct average treatment effect bounds as in  \citeauthor{lee2009training} (\citeyear{lee2009training}). For these bounds to be valid, a monotonicity assumption must hold on selection. That is, treatment can only affect selection in one direction. In my case, that means that EHR exposure increases the likelihood of exiting or moving workplace settings and no hospitalists remain in the sample under treatment who would have exited or moved in the absence of EHR exposure. Since I find in earlier analyses that hospitalists are more likely to move and exit due to EHRs, this assumption is reasonable. To compute the bounds, I trim the distribution of patient count in the control group by the difference in attrition rate. Trimming the bottom (top) and taking the differences in expectation between treatment and trimmed control provides a lower (upper) bound. This bounding yields an interval estimate of 54.6 patients to 125.4 patients, compared to the main result of 28 patients. The bound therefore suggests selection of physicians pushes the estimate down.

Similarly, I investigate whether this increase in the number of patients seen is due to the decrease in the number of working hospitalists. Specifically, a hospital that loses a portion of their physicians, and divides those patients among those remaining. In the long run, the hospital should hire new hospitalists to replace those that left, but in the short run estimates could be driven by increased workload instead of EHR use. This could explain why there is a much larger increase in the first year after exposure compared to the magnitudes of the estimates as they level out over time. Using the estimates I found in Section \ref{sec:retire}, I calculate the increase in the number of patients that would occur if all patients from hospitalists who exited were divided between remaining hospitalists. In the year of exposure the entire increase in patient count is due to EHR use since I find no effect on the likelihood of exit. In the first year after exposure, I estimate the increase in patients due to exiting physicians to be approximately 20 patients, which is less than the observed patient count estimates presented above. However, the second year after exposure is when I observe that largest effect on exiting, and so I do not rule out that the increase in patient count for every year after the first could be caused by the decrease in hospitalists.

Finally, I consider whether EHR exposure affects claims filed per patient. On average, hospitalists file roughly 4 claims per patient. I present estimates in Figure \ref{fig:claima}, which shows that EHRs tend to increase claims filed per patient over time. The magnitude of the effect ranges from .2 to .8, or 5-20\% relative to the mean, indicating that as hospitalists are exposed to the new software, they bill for more items on average.  I also show heterogeneity of this result by age group, and Figure \ref{fig:claimb} shows that younger hospitalists are exhibiting more of an increase in claim count than older hospitalists. However, when I analyze this relationship under different data specifications (shown in Appendix \ref{sec:chart}, a majority of the estimates are negative or zero. Anticipation and limiting sample years cause the effect to disappear or even become negative. Thus, it is not clear whether this positive effect is actually an unintended consequence of EHRs and I do not make any strong conclusions.

@@

@@

There are various implications of these findings. First, many hospitalists left clinical settings in a concentrated number of years due to EHR exposure, potentially worsening patient access to care. Second, in a field where younger physicians learn many skills through observing senior physicians, this particular loss of senior hospitalists may have long term negative effects on patient outcomes, but this is a question left to future research. Third, hospitalists are exhibiting behavior that suggests they incur switching costs to avoid external burdens in the workplace. When policymakers consider regulations on physician practice it is important to understand that physicians may take measures to avoid such regulation. Finally, these results speak to the ongoing debate of whether EHRs make health care more efficient. My findings suggest that EHRs make hospitalists more productive through the number of patients they see. However, the limitations of the data prevent a solid understanding of the amount of time spent with patients, as there may be a trade-off between comprehensive patient care and productivity. This question is left to future research.

@@

For years 2009-2015, the CMS collected detailed information on the number of Medicare patients shared between any two NPIs within 30, 90, or 180 day intervals.\footnote{Data can be dowloaded from \hyperlink{https://www.nber.org/research/data/physician-shared-patient-patterns-data}{https://www.nber.org/research/data/physician-shared-patient-patterns-data}} I use data that captures shared patient activity in 30 day intervals. I merge the shared patient data to the filtered tax code data created in Section \ref{sec:taxcode} to identify NPIs who are either PCPs or hospitals. I filter the pairs to include one physician and one hospital; there are 12.6 million of these pairs. Some pairs are duplicates from the hospital being listed first or second in the shared patient data (being listed first means the patient went to NPI 1 first, but my analysis does not depend on the order of care, only the relationship between hospital and physician as a whole). I combine duplicates into one observation, summing the same day count variable. Once duplicates are removed, there are 7.1 million observations. Most of the pairs have very few shared patients, which is not indicative of the physician working inside the hospital. I drop any pairs who do not have at least 30 same-day shared patients per year the pair appears in the data.\footnote{As a sensitivity analysis, I also consider thresholds of 10 and 60. Results are similar and can be found in Figure \ref{sec:chart}.} The final list of pairs consists of 1.3 million observations.

@@

I combine the physician-hospital pairs created in Section \ref{sec:sharedpat} with various data sets for hospital or physician level information. First, I merge to CMS Physician Compare\footnote{Data can be downloaded from \hyperlink{https://data.cms.gov/provider-data/dataset/mj5m-pzi6}{https://data.cms.gov/provider-data/dataset/mj5m-pzi6}} for information on each physician's graduation date. This data contains more information on physician quality, but I am limited to time-invariant information since Physician Compare spans 2012-2015 and 2010-2012 is a time of major EHR implementation. I drop any physicians who graduated medical school after 2004, since graduating after that means leaving residency or graduating during the span of my main data, 2009-2017, and potentially exhibiting labor market changes that seem associated with EHRs but are not.

@@

Next, I create the dependent variables used in the analysis. The first outcome is whether a physician chooses to exit work in a clinical setting. First, I sum a physicians claim counts across zip codes into one variable for total claim count in a given year. Then, I create a variable that sums up a physicians claims in all future years. In 2009, this variable sums up all claims from 2010-2017, and so on. Then I create a variable for the first year that a physician has a future claim count of zero. However, this counts exit one year too early. For example, if a physician has no claims from 2014 to 2017, the minimum year variable is set to 2013, but the first year of exit is 2014. Thus, I add one year to the minimum year variable. I then define the outcome variable, exiting, as one if a hospitalist ever exits and the year is equal to the first year of exit. I also create a physician level variable for whether they ever retire for sample limitations in analyzing other outcomes.

@@

As referenced in the main paper, the effect of EHR exposure on patient count could be biased under any demand-driven events happening at the same time as EHR adoption in hospitals. The only nation-wide demand altering event that I know of during the time of this sample is the Affordable Care Act, which took effect mainly in 2014. If this is driving my positive finding, this would show up in a de-aggregated treatment effects by year of exposure. In Figure \ref{fig:patientgroup}, I show that the effects persist in all years of exposure, not just in 2014.

@@

\subsection{Billing Activity as Cause for Work Setting Shift}

@@

There is a robust recent literature considering the issues with estimating heterogeneous treatment effects with a two way fixed effects specification, the classic approach in staggered treatment setups. To avoid these problems, in my main specification I use the average group time effect estimator established in \citeauthor{callaway2021difference} (\citeyear{callaway2021difference}). In this section, I discuss other potential estimators and why I ultimately included average group time treatment effects as the main specification.

@@

One method that addresses issues with the TWFE specification by differencing out fixed effects is established in \citeauthor{gardner2021two} (\citeyear{gardner2021two}). This approach is commonly known as two stage difference-in-differences. The approach requires a group of never-treated units as the comparison group, a drawback for my study in that I remove all never-treated units to utilize all years of MD-PPAS data, and because never-treated hospitals are likely compositionally different from those treated in the majority of the sample. Nevertheless, I limit the years of data to 2009-2015 and present estimates as a percentage of the mean of each variable in Figure \ref{fig:estimators}. Generally, the results have the same sign with a slightly larger magnitude.

@@

>>>>

Two key assumptions underlie the main analysis. First, I assume that hospitalists are constrained to use the technology implemented in their work setting. Second, I assume the decision for hospitals to implement EHRs is exogenous to individual hospitalists, and uncorrelated with other major changes to the hospitalist work environment. I mitigate concerns of the latter by analyzing whether other observable hospital changes are correlated with EHR adoption, and find no evidence that this could be biasing the results. Additionally, the main findings are not sensitive to a plethora of robustness checks. The results suggest that a surge of hospitalists left clinical settings altogether, and even more that shifted work away from the hospitals where the EHR was deployed and towards outpatient/office settings. Thus, EHR implementation may have intensified access to care issues for hospitalized patients. This unintended consequence of incentivizing rapid technology adoption highlights the need for careful consideration of technologies and policies that might negatively affect work environment.

@@

The Medicare Data on Provider Practice and Specialty (MD-PPAS) is a database of physicians which details claim counts, physician specialty, and practice location from 2009-2017. Variables included are physician NPI, primary and secondary specialties, patient count and total billing counts, and fraction of patients seen in different settings (hospital, office/outpatient, emergency room, etc.). I only include physicians classified as hospitalists\footnote{Physician who self-identifies as hospitalist or has 70\% of patients in inpatient setting. I cannot distinguish between types of hospitalists as the data source manipulates this definition based on percent of patients seen in a hospital, so these can be formal hospitalists or internists.}, internal medicine, general practice or family practice in at least one year. I drop physicians with less than 70\% of patients in a hospital setting in all years, as the focus is on those who are affected by a hospital's decision to adopt (I investigate robustness to this threshold in Appendix \ref{sec:changes}). I use this data to construct the dependent variables in my analysis.

@@

I aggregate the data to the hospitalist level. I drop those who are not treated by 2015 since I do not observe whether they become treated in 2016 or 2017.\footnote{In Appendix \ref{sec:changes} I limit years to 2009-2015 and include never-treated. The results are robust to this specification choice.} That is, the data contains no units which are ``never-treated". Finally, I include a physician's graduation year from Physician Compare (provided by CMS). I limit to hospitalists who graduated before 2005, as anyone who graduated medical school after that will be finishing residency during the span of the data and will exhibit labor market changing behavior which could be correlated to EHR exposure purely because of switching work settings during a time of rapid EHR implementation.There are various thresholds used in the construction of this final data, including how many shared patients constitutes a physician-hospital relationship. The details of such thresholds are explained in Appendix \ref{app:data}. To ensure that the results  are not sensitive to various choices of the thresholds, I estimate the effect of EHR exposure on all outcomes under different combinations of data limitations. The specification charts with these results are shown in Appendix \ref{sec:chart}, and I discuss each in conjunction with the main results.

@@

As discussed in Section \ref{sec:worksetting}, hospitalists could be switching work settings either because of their own preferences to avoid technology, or because EHRs, through improved substitution between work settings, cause hospitals to re-allocate workers to hospital-owned outpatient settings. Due to lack of granular EHR data, it is not clear whether acquired outpatient facility EHRs are interoperable with inpatient EHRs, even if the two have the same EHR vendor. Outpatient EHRs are designed to be much more simplistic than inpatient EHRs which communicate across departments. The main purpose of an outpatient EHR is to track patient data, but it is unlikely this tracking covers the entire health system in the years covered by the data, before interoperability initiatives began. Still, it is theoretically possible that EHRs facilitate easier substitution between facilities. Therefore, I investigate this mechanism through a heterogeneity analysis of hospitals with different contract types with physicians, which will shut down certain channels of re-allocation of resources.

I do not observe individual physician contract type with each hospital, but I do observe hospital level information about the aggregate nature of their physician contracts. To investigate the potential mechanism of hospital allocation, I present a heterogeneity analysis for all outcomes broken down by level of integration the physician is exposed to in their connected hospitals. From least integrated (contracts with multiple facilities) to most integrated (full salaried hospitalists), results are broken down by Independent Practice Association (IPA), Open Physician-Hospital Organization (OPHO), Closed Physician-Hospital Organization (CPHO), and Integrated Salary Model (ISM). In practice, hospitals drive the re-allocation of work setting only if if the hospitalist is heavily integrated. For contracted hospitalists, the choice remains whether to contract with a hospital who wants them to work in an outpatient setting. However, estimates do not differ between the types of integration relationships, indicating that the results are not driven entirely by hospital-owned outpatient facilities. I present summary statistics by contract type and estimates of this heterogeneity analysis for all outcomes in Appendix

\begin{figure}
    \centering
    \caption{Effect of EHR Exposure on Outcomes by Integration Level}
    \includegraphics[width=.7\textwidth]{Objects/heterog_plot_int_office.pdf}
    \label{fig:het_int}
\end{figure}

Another proxy for contract type is whether the hospitalists works with one system or is connected to facilities from different systems. I present the results from this heterogeneity analysis in Figure \ref{fig:heter}, and find no difference in effects between the two groups.

Finally, I investigate whether this finding is potentially due EHRs changing the type of billing performed by hospitalists. For example, they could be performing the same services in the same location, but coding the billing information differently. I do not find this to be the case. This analysis can be found in Appendix \ref{app:coding}.

@@

While I am investigating the supply side factors of this outcome, patient count is also driven by demand for health care. A major change in the U.S. health care system, the Affordable Care Act (ACA) was passed in 2010, and many changes were made that could affect the demand for health care, such as insurance expansions, and hospital quality initiatives. In order for additional changes to be biasing my results, they must be taking place in such a way that they either cause EHR adoption in hospitals or are differentially affecting EHR exposed hospitalists vs. non-EHR exposed hospitalists. I cannot completely rule this out, but it seems unlikely that any one change is causing a large and immediate effect for every treated group year. If the results are driven primarily by the group of hospitals who implemented EHRs in 2012 (when pay-for-performance initiatives began) or 2014 (when Medicaid expansion went into effect), one might be concerned that the legislation is driving the increased patient count. The ACA had potentially opposing effects on Medicare patients: they may be crowded out by an increased demand in Medicaid patients, or the small changes to Medicare may have led to an increase in utilization from the Medicare population. A recent study found no negative spillovers to the Medicare population due to the ACA (\cite{carey2020impact}), suggesting that this particular change should not drive the results. Additionally, I present average group time treatment effects for each treated group in Figure \ref{fig:patientgroup} of the appendix. Despite different EHR exposure years, each group shows the same pattern of an increase in patient count after exposure, suggesting that the groups most affected by the ACA are not driving the main findings.

EHR exposure affects which hospitalists are included in this analysis (since I drop those who exit or move), indicating a non-random attrition problem. That is, if productive or tech savvy hospitalists are the ones who remain in the same hospital after EHR exposure, the estimates are biased. The proportion of those who are not included in this analysis is 18\% in the control group and 36\% in the treated group. To investigate whether these results are robust to this issue, I construct average treatment effect bounds as in  \citeauthor{lee2009training} (\citeyear{lee2009training}). For these bounds to be valid, a monotonicity assumption must hold on selection. That is, treatment can only affect selection in one direction. In this case, that means that EHR exposure increases the likelihood of exiting or moving workplace settings and no hospitalists remain in the sample under treatment who would have exited or moved in the absence of EHR exposure. Since I find in earlier analyses that hospitalists are more likely to move and exit due to EHRs, this assumption is reasonable. To compute the bounds, I trim the distribution of patient count in the control group by the difference in attrition rate. Trimming the bottom (top) and taking the differences in expectation between treatment and trimmed control provides a lower (upper) bound. This bounding yields an interval estimate of 54.6 patients to 125.4 patients, compared to the main result of 28 patients. The bound therefore suggests selection causes a downward bias in the estimated effect, but the implications of the findings remain.

Similarly, I investigate whether this increase in the number of patients seen is due to the decrease in the number of working hospitalists. Specifically, a hospital that loses a portion of their physicians, and divides those patients among those remaining. In the long run, the hospital should hire new hospitalists to replace those that left, but in the short run estimates could be driven by increased workload instead of EHR use. This could explain why there is a much larger increase in the first year after exposure compared to the magnitudes of the estimates as they level out over time. Using the estimates I found in Section \ref{sec:retire}, I calculate the increase in the number of patients that would occur if all patients from hospitalists who exited were divided between remaining hospitalists. In the year of exposure the entire increase in patient count is due to EHR use since I find no effect on the likelihood of exit. In the first year after exposure, I estimate the increase in patients due to exiting physicians to be approximately 20 patients, which is less than the observed patient count estimates presented above. However, the second year after exposure is when I observe that largest effect on exiting, and so I do not rule out that the increase in patient count for every year after the first could be caused by the decrease in hospitalists. Further, I find that EHR adoption is not correlated with the aggregate number of physicians working in the hospital, shown in Table \ref{tab:hosp_assoc}, suggesting that hospitals replace those who exit relatively quickly.

Finally, I consider whether EHR exposure affects claims filed per patient. On average, hospitalists file roughly 4 claims per patient. I present estimates in Figure \ref{fig:claima}, which shows that EHRs tend to increase claims filed per patient over time. The magnitude of the effect ranges from .2 to .8, or 5-20\% relative to the mean, indicating that as hospitalists are exposed to the new software, they bill for more items on average.  I also show heterogeneity of this result by age group, and Figure \ref{fig:claimb} shows that younger hospitalists are exhibiting more of an increase in claim count than older hospitalists. However, when I analyze this relationship under different data specifications, shown in Appendix \ref{sec:chart}, a majority of the estimates are negative or zero. Anticipation and limiting sample years cause the effect to disappear or even become negative. Thus, it is not clear whether this positive effect is actually an unintended consequence of EHRs, and I do not make any strong conclusions.

@@

One concern is that EHR adoption could be occurring simultaneously with a host of other hospital initiatives to improve productivity or quality, which may also affect physician work environment. If true, EHR adoption is not exogenous and the results found above are biased. Here, I discuss the implications of this type of endogeneity for each outcome, and present evidence that hospital behaviors are not changing in a way that is correlated with EHR adoption.

Anecdotally, there is no other change made in hospitals during this time period that caused the frustration and burn out that EHRs did. Thus, on average, it is unlikely that other hospital level changes are imposing a burden on hospitalist work environment that could explain physician exit or work setting changes. If work environment changes are occurring, it is more likely that they offset some of the cost imposed by EHRs, and that my estimates for exit and work setting are a lower bound of the true effect of EHRs.

Therefore, this endogeneity is more of a concern when considering productivity of hospitalists. If other work environment factors correlated with EHR adoption improve productivity, then the increase I find in productivity could actually be due to other factors. I investigate this endogeneity by constructing a hospital level data set from 2009 to 2015 with various hospital-level outcomes that EHR adoption may be correlated with. While not all hospital behaviors are observable, it is very likely that if the hospital is making a large portfolio of changes along with EHR, we will see a correlation between EHR adoption and observed hospital behaviors. I categorize the behaviors into two categories: non-EHR investments and organizational changes, and level of hospital integration with physicians. Non-EHR investments and organization changes include building costs (ex. did the hospital renovate or expand?), equipment costs (ex. did the hospital invest in other machines?), investment in subsidiaries, changes in the main hospital administrator, and number of beds. Relationships with physicians include the number of physicians employed, and indicators for the types of vertical integration with physicians: independent practice organization (IPA), open physician-hospital organization (OPHO), closed physician-hospital organization (CPHO), and fully integrated organization (FIO), listed from the lowest level of integration to the highest. An IPA has loose contracts with physicians, but the physicians remain independent, an OPHO adds administration tasks to the physician, a CPHO contracts with specific physicians, and a FIO hires physicians as salaried employees (\cite{short2017integration}). The expense variables are pulled from the Healthcare Cost Reporting Information System (HCRIS) database, and the other hospital characteristics come from the AHA data.

\import{Table Code}{hosp_analysis.tex}

I estimate the effect of EHR adoption on each of these outcomes using the same \citeauthor{callaway2021difference} (\citeyear{callaway2021difference}) staggered difference-in-differences methodology. I present coefficients and standard errors from this analysis in Table \ref{tab:hosp_assoc}. Again, I do not present these to suggest that EHR may have a causal impact on other behaviors, but to observe whether the behaviors seem to be happening in conjunction with EHR adoption. Based on the confidence intervals, none of the behaviors considered seem to be occurring simultaneously with EHR adoption on average. While I cannot rule out all changes that could be occurring at the hospital level, the lack of movement in other important behaviors mitigates the concern.

@@

There are various implications of these findings. First, many hospitalists left clinical settings in a concentrated number of years due to EHR exposure, potentially worsening patient access to care. Second, in a field where younger physicians learn many skills through observing senior physicians, this particular loss of senior hospitalists may have long term negative effects on patient outcomes, but this is a question left to future research. Third, hospitalists are exhibiting behavior that suggests they incur switching costs to avoid external burdens in the workplace. When policymakers consider regulations on physician practice it is important to understand that physicians may take measures to avoid such regulation. Finally, these results speak to the ongoing debate of whether EHRs make health care more efficient. These findings suggest that EHRs make hospitalists more productive through the number of patients they see. However, the limitations of the data prevent a solid understanding of the amount of time spent with patients, as there may be a trade-off between comprehensive patient care and productivity. This question is left to future research.

@@

For years 2009-2015, the CMS collected detailed information on the number of Medicare patients shared between any two NPIs within 30, 90, or 180 day intervals.\footnote{Data can be dowloaded from \hyperlink{https://www.nber.org/research/data/physician-shared-patient-patterns-data}{https://www.nber.org/research/data/physician-shared-patient-patterns-data}} I use data that captures shared patient activity in 30 day intervals. I merge the shared patient data to the filtered tax code data created in Section \ref{sec:taxcode} to identify NPIs who are either PCPs or hospitals. I filter the pairs to include one physician and one hospital; there are 12.6 million of these pairs. Some pairs are duplicates from the hospital being listed first or second in the shared patient data (being listed first means the patient went to NPI 1 first, but the analysis does not depend on the order of care, only the relationship between hospital and physician as a whole). I combine duplicates into one observation, summing the same day count variable. Once duplicates are removed, there are 7.1 million observations. Most of the pairs have very few shared patients, which is not indicative of the physician working inside the hospital. I drop any pairs who do not have at least 30 same-day shared patients per year the pair appears in the data.\footnote{As a sensitivity analysis, I also consider thresholds of 10 and 60. Results are similar and can be found in Figure \ref{sec:chart}.} The final list of pairs consists of 1.3 million observations.

@@

I combine the physician-hospital pairs created in Section \ref{sec:sharedpat} with various data sets for hospital or physician level information. First, I merge to CMS Physician Compare\footnote{Data can be downloaded from \hyperlink{https://data.cms.gov/provider-data/dataset/mj5m-pzi6}{https://data.cms.gov/provider-data/dataset/mj5m-pzi6}} for information on each physician's graduation date. This data contains more information on physician quality, but I am limited to time-invariant information since Physician Compare spans 2012-2015 and 2010-2012 is a time of major EHR implementation. I drop any physicians who graduated medical school after 2004, since graduating after that means leaving residency or graduating during the span of the main data, 2009-2017, and potentially exhibiting labor market changes that seem associated with EHRs but are not.

@@

Next, I create the dependent variables used in the analysis. The first outcome is whether a physician chooses to exit work in a clinical setting. First, I sum a physicians claim counts across zip codes into one variable for total claim count in a given year. Then, I create a variable that sums up a physicians claims in all future years. In 2009, this variable sums up all claims from 2010-2017, and so on. Then I create a variable for the first year that a physician has a future claim count of zero. However, this counts exit one year too early. For example, if a physician has no claims from 2014 to 2017, the minimum year variable is set to 2013, but the first year of exit is 2014. Thus, I add one year to the minimum year variable. I then define the outcome variable, exiting, as one if a hospitalist ever exits and the year is equal to the first year of exit. I also create a physician level variable for whether they ever retire for sample limitations in analyzing other outcomes. Since exit is relatively rare, I present a distribution showing where they occur geographically. The distribution is fairly uniform.

\begin{figure}[ht!]
    \centering
    \caption{Distribution of Exiters}
    \includegraphics[width=\textwidth]{Objects/retire_dist_map.pdf}
    \label{fig:retire_dist}
\end{figure}

@@

As referenced in the main paper, the effect of EHR exposure on patient count could be biased under any demand-driven events happening at the same time as EHR adoption in hospitals. The only nation-wide demand altering event that I know of during the time of this sample is the Affordable Care Act, which took effect mainly in 2014. If this is driving the positive finding, this would show up in a de-aggregated treatment effects by year of exposure. In Figure \ref{fig:patientgroup}, I show that the effects persist in all years of exposure, not just in 2014.

@@

\subsection{Coding Activity as Cause for Work Setting Shift}\label{app:coding}

@@

There is a robust recent literature considering the issues with estimating heterogeneous treatment effects with a two way fixed effects specification, the classic approach in staggered treatment setups. To avoid these problems, in the main specification I use the average group time effect estimator established in \citeauthor{callaway2021difference} (\citeyear{callaway2021difference}). In this section, I discuss other potential estimators and why I ultimately included average group time treatment effects as the main specification.

@@

One method that addresses issues with the TWFE specification by differencing out fixed effects is established in \citeauthor{gardner2021two} (\citeyear{gardner2021two}). This approach is commonly known as two stage difference-in-differences. The approach requires a group of never-treated units as the comparison group, a drawback of this study in that I remove all never-treated units to utilize all years of MD-PPAS data, and because never-treated hospitals are likely compositionally different from those treated in the majority of the sample. Nevertheless, I limit the years of data to 2009-2015 and present estimates as a percentage of the mean of each variable in Figure \ref{fig:estimators}. Generally, the results have the same sign with a slightly larger magnitude.

@@

\import{Table Code}{int_sumstats.tex}

\begin{figure}
    \centering
    \caption{Heterogeneity Analysis, Hospital Contract Types}
    \includegraphics[scale=.6]{Objects/heterog_plot_int.pdf}
\end{figure}
